---
title: "Sporadic colorectal cancer SKAT"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

**Started**: Alexey Larionov, 14Jul2020  
**Last updated**: Rofaida Desoki, 17Jul20

# Summary  

Sporadic colorectal cancer: SKAT burden and variance-aggregation tests  

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r}

options(width = 999)

```

# Start section

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Additional options
options(stringsAsFactors = F)

# Working folders
base_folder <- "/home/rdesoki/re_gecip/inherited_cancer_predisposition/rofaida/Rproject"

scripts_folder <- file.path(base_folder,"scripts","s09_sporadic_colon_cancer")
setwd(scripts_folder)

data_folder <- file.path(base_folder,"data","s09_sporadic_colon_cancer")

plink_folder <- file.path(data_folder,"plink")
covars_sets_folder <- file.path(data_folder,"covars_sets")
ssd_folder <- file.path(data_folder,"ssd")

# Required library
library(SKAT)

```

# Read SSD.INFO  

```{r}

# Source files
File.SSD = file.path(ssd_folder,"dataset.ssd")
File.Info = file.path(ssd_folder,"dataset.info")

# Read SSD.INFO
SSD.INFO <-Open_SSD(File.SSD, File.Info)

# Explore SSD.INFO
SSD.INFO

# Clean-up
rm(File.SSD, File.Info)

```

# Read fam and cov

```{r}

File.Fam = file.path(plink_folder,"dataset.fam")
File.Cov = file.path(covars_sets_folder,"dataset.cov")

fam_cov.df <- Read_Plink_FAM_Cov(File.Fam, File.Cov, Is.binary=TRUE, cov_header=TRUE)

dim(fam_cov.df)
head(fam_cov.df)
table(fam_cov.df$Phenotype)

rm(File.Fam, File.Cov)

```

# Null model

Becomes large (~4GB) with 10^5 resampling

```{r}

obj<-SKAT_Null_Model(Phenotype ~ sex + age + pc1 + pc2 + pc3 + pc4 + pc5 + pc6, 
                     data=fam_cov.df, out_type="D")
                     #, n.Resampling=10000)

rm(fam_cov.df)

```

# Burden test

```{r}

# Calculate p-values using the dafault Hybrid approach
out.burden <- SKATBinary.SSD.All(SSD.INFO, obj, method="Burden")
#warnings()

# Explore out.burden object
attributes(out.burden)

# Extract data frame with results
burden.df <- out.burden$results

# Add multiple-testing correction
p_adj <- p.adjust(burden.df$P.value, method="fdr")
burden.df <- data.frame(burden.df,p_adj)
burden.df

# Genes passing Family-Wise-Error-Rate 0.05 from resampling
# Resampling_FWER(out.burden,FWER=0.05)

# Number of tests that might be used for Bonferroni correction
# Get_EffectiveNumberTest(out.burden$results$MAP, alpha=0.05)

# MAP-adjusted QQ-plot
QQPlot_Adj(out.burden$results$P.value, out.burden$results$MAP,
           main="Sporadic Colon Cancer\nBurden test", Is.unadjsted = F)

# Clean-up
rm(p_adj)

```

# Variance-aggregation test

```{r}

out.skat<-SKATBinary.SSD.All(SSD.INFO, obj, method="SKAT")
#warnings()

# Explore out.burden object
attributes(out.skat)

# Extract data frame with results
skat.df <- out.skat$results

# Add multiple-testing correction
p_adj <- p.adjust(skat.df$P.value, method="fdr")
skat.df <- data.frame(skat.df,p_adj)
skat.df

# Genes passing Ffamily-Wise-Eerror-Rate 0.05 from resampling
# Resampling_FWER(out.skat,FWER=0.05)

# Number of tests that might be used for Bonferroni correction
# Get_EffectiveNumberTest(out.skat$results$MAP, alpha=0.05)

# MAP-adjusted QQ-plot
QQPlot_Adj(out.skat$results$P.value, out.skat$results$MAP,
           main="Sporadic Colon Cancer\nVaraince-aggregation test", Is.unadjsted = F)

# Clean-up
rm(obj, SSD.INFO, p_adj)

```

# Write results to text files

```{r}

# Write results of burden tests
file_name <- file.path(data_folder,"s05_burden.txt")
write.table(out.burden$results, file = file_name, sep="\t", 
            quote = FALSE, row.names = FALSE, col.names = TRUE)

# Write results of variance-aggregation tests
file_name <- file.path(data_folder,"s05_skat.txt")
write.table(out.skat$results, file = file_name, sep="\t",
            quote = FALSE, row.names = FALSE, col.names = TRUE)

# Clean-up
rm(file_name)

```

# Close SSD

Recomended by SKAT

```{r}

Close_SSD()

```

Save results

```{r}

save.image(file.path(data_folder,"s05_SKAT.RData"))

```

# End section 

```{r}

ls()
sessionInfo()
Sys.time()

```
