---
title: "Explore Sporadic Cancers"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

**Last updated**: Rofaida, 15Mar2020  

# Summary  

Explore Sporadic Cancers

# Start section

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r}
options(width=999)
```

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Dont use factors by default
options(stringsAsFactors = F)

# Folders
base_folder <- "/home/rdesoki/re_gecip/inherited_cancer_predisposition/rofaida/Rproject"

# Libraries
library(dplyr)

```

# load data  

```{r}

load(paste(base_folder,"data","s01_import_Labkey_tables","s02_clean_Labkey_tables.RData", sep="/"))

scripts_folder <- paste(base_folder,"scripts","s02_explore_Labkey_data",sep="/")

rm(data_folder)

ls()

dim(vcf_participants.df)
dim(sporadic_cancers.df)
dim(rare_diseases.df)

```

# Sporadic Cancers table  

```{r}

str(sporadic_cancers.df)

# Disease type (note how to pring all rows in a tible)
sporadic_cancers.df %>% 
  group_by(cancer_disease_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts)) %>% 
  print(n=Inf)

# Disease sub-type
sporadic_cancers.df %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts)) 

# Disease sub-type
sporadic_cancers.df %>% 
  group_by(disease) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# No missed data
sum(is.na(sporadic_cancers.df$disease))

```

# Specific cancers -- with subtypes

```{r}

# Breast
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "BREAST") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Colorectal
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "COLORECTAL") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Lung
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "LUNG") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Renal
sporadic_cancers.df %>%
  filter(cancer_disease_type == "RENAL") %>%
  group_by(cancer_disease_sub_type) %>%
  summarise(counts=n()) %>%
  arrange(desc(counts))

# Sarcomas
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "SARCOMA") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Eendometrial
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "ENDOMETRIAL_CARCINOMA") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Blood
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "HAEMONC") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Ovarian
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "OVARIAN") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Prostate
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "PROSTATE") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Adult glioma
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "ADULT_GLIOMA") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Bladder
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "BLADDER") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Hepato-Pancreato-Bilary
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "HEPATOPANCREATOBILIARY") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Malignant Melanoma
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "MALIGNANT_MELANOMA") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Oral OroPharingeal
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "ORAL_OROPHARYNGEAL") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Upper gastrointestinal
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "UPPER_GASTROINTESTINAL") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Pediatric
sporadic_cancers.df %>% 
  filter(cancer_disease_type == "CHILDHOOD") %>% 
  group_by(cancer_disease_sub_type) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

```

# Participants with multiple cancers    

After consulting with MT it has been decided to keep all records if a participant has two (or more) cancers because such participants are more likely to have cancer predisposition variants.  Such participants should be included into analysis for each of the involved cancers.  Howevewr, a care should be taken that they are not included multiple times when it is not appropriate (upon aggregating genes).  

```{r}

multi_cancer_ids.df <- sporadic_cancers.df %>% 
  filter(duplicated(sporadic_cancers.df$participant_id)) %>% 
  select(participant_id)

multi_cancers.df <- sporadic_cancers.df %>% 
  filter(participant_id %in% multi_cancer_ids.df$participant_id) %>% 
  select(participant_id, disease)

dim(multi_cancers.df)
length(unique(multi_cancers.df$participant_id))
head(multi_cancers.df, n=20)

rm(multi_cancer_ids.df, multi_cancers.df)

```

# Save  result

```{r}

# No changes to save

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()

```
