---
title: "Explore controls"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

**Last updated**: Rofaida, 17Mar20  

# Summary  

Select and count **controls**  
Unaffected parents of probands with intellectual disorders  

# Start section

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r}
options(width=999)
```

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Dont use factors by default
options(stringsAsFactors = F)

# Folders
base_folder <- "/home/rdesoki/re_gecip/inherited_cancer_predisposition/rofaida/Rproject"

# Libraries
library(dplyr)

```

# load data  

```{r}

load(paste(base_folder,"data","s01_import_Labkey_tables","s02_clean_Labkey_tables.RData", sep="/"))

scripts_folder <- paste(base_folder,"scripts","s02_explore_LabKey_data",sep="/")

data_folder <- paste(base_folder,"data","s02_explore_Labkey_data",sep="/")
dir.create(data_folder,showWarnings=F)

ls()

dim(vcf_participants.df)
dim(sporadic_cancers.df)
dim(rare_diseases.df)

```

# Select participants with Intellectual Disability Disorders (idd)  

Note one duplicated participant_id: its OK, because probands may have several disorders  

```{r}

# Participant_id-s for probands with Intellectual Disability Disorders from Rare Ddiseases
intellectual_disability.df <- rare_diseases.df %>% 
  filter(normalised_specific_disease=="Intellectual disability") %>% 
  select(participant_id)

dim(intellectual_disability.df)
sum(duplicated(intellectual_disability.df$participant_id))

# Family IDs and parents data for the probands with Intellectual Disability Disorders
intellectual_disability_family.df <- vcf_participants.df %>% 
  filter(participant_id %in% intellectual_disability.df$participant_id) %>% 
  select(participant_id, participant_type, father_affected, mother_affected,rare_diseases_family_sk)

# Explore table with familly IDs and parents data
dim(intellectual_disability_family.df)

sum(is.na(intellectual_disability_family.df$participant_type))
table(intellectual_disability_family.df$participant_type)

sum(is.na(intellectual_disability_family.df$father_affected))
table(intellectual_disability_family.df$father_affected)

sum(is.na(intellectual_disability_family.df$mother_affected))
table(intellectual_disability_family.df$mother_affected)

# Clean-up
rm(intellectual_disability.df)

```

# Select uaffected parents of IDD probands  

### Fathers  

```{r}

# Select family IDs of unaffected fathers
fathers_families.df <- intellectual_disability_family.df %>% 
  filter(participant_type == "Proband",
         father_affected == "No") %>% 
  select(rare_diseases_family_sk)

dim(fathers_families.df)
sum(duplicated(fathers_families.df$rare_diseases_family_sk))

# Select fathers ID-s from the families
fathers_id.df <- vcf_participants.df %>% 
  filter(rare_diseases_family_sk %in% fathers_families.df$rare_diseases_family_sk,
         biological_relationship_to_proband == "Father") %>% 
  select(participant_id)

dim(fathers_id.df)
sum(duplicated(fathers_id.df$participant_id))

# Clean-up
rm(fathers_families.df)

```

### Mothers  

```{r}

# Select family IDs of unaffected  mothers
mothers_families.df <- intellectual_disability_family.df %>% 
  filter(participant_type == "Proband",
         mother_affected == "No") %>% 
  select(rare_diseases_family_sk)

dim(mothers_families.df)
sum(duplicated(mothers_families.df$rare_diseases_family_sk))

# Select  mothers ID-s from the families
mothers_id.df <- vcf_participants.df %>% 
  filter(rare_diseases_family_sk %in% mothers_families.df$rare_diseases_family_sk,
         biological_relationship_to_proband == "Mother") %>% 
  select(participant_id)

dim( mothers_id.df)
sum(duplicated(mothers_id.df$participant_id))

# Clean-up
rm(mothers_families.df, intellectual_disability_family.df)

```

### Merge fathers and mothers ID-s

```{r}

unaffected_parents_idd.df <- vcf_participants.df %>% 
  filter(participant_id %in% mothers_id.df$participant_id | participant_id %in% fathers_id.df$participant_id) %>% 
  select(participant_id, biological_relationship_to_proband, age_at_consent)

dim(unaffected_parents_idd.df)

rm(fathers_id.df, mothers_id.df)

```

# Explore selected controls  

Despite selecting the non-affected parents of idd-probands, some of them still have idd and some other rare disease(s).  These controls are removed.  

```{r}

sum(duplicated(unaffected_parents_idd.df$participant_id))
table(unaffected_parents_idd.df$biological_relationship_to_proband)
hist(unaffected_parents_idd.df$age, lab=T, ylim=c(0,1200))

controls_id.df <- unaffected_parents_idd.df %>% 
  select(participant_id)

dim(controls_id.df)

rare_diseases.df %>% 
  filter(participant_id %in% controls_id.df$participant_id) %>% 
  group_by(normalised_disease_group) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

rare_diseases.df %>% 
  filter(participant_id %in% controls_id.df$participant_id) %>% 
  arrange(normalised_disease_group, normalised_disease_sub_group)

affected_controls.df <- rare_diseases.df %>% 
  filter(participant_id %in% controls_id.df$participant_id) %>% 
  select(participant_id)

dim(affected_controls.df)

unaffected_parents_idd.df <- unaffected_parents_idd.df %>% 
  filter(!participant_id %in% affected_controls.df$participant_id)

dim(unaffected_parents_idd.df)

rm(affected_controls.df, controls_id.df)

```

# Save  result

```{r}

save.image(paste(data_folder,"s04_explore_controls.RData",sep="/"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()

```