---
title: "Read VCF with vcfR"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

**Started**: Alexey Larionov, 10 July 2020  
**last updated**: Rofaida Desoki, 13 July 2020

# Summary

Add phenotype data to the dataset  

## Note about duplicated plate-keys

Phenotype data are NOT synchronized with genotypes at this stage because there are 88 duplicated plate-keys phenotypes.  These are participants with multiple tumours: i.e. there could be several phenotypes (=diagnoses) for the same platelkey.  

However, there should be no duplicated platekeys within a single diagnosis.  So, the genotype and phenotype data will be synchronized within the sub-groups immediately before SKAT analysis.  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}

#options(width=999)

```

# Start section

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Working folders
base_folder <- "/home/rdesoki/re_gecip/inherited_cancer_predisposition/rofaida/Rproject"

```

# Read phenotype data

```{r}

load(file.path(base_folder,"data","s03_select_sample","s04_check_relatedness.RData"))

ls()

dim(participants.df)
colnames(participants.df)
participants.df[1:5,1:5]

```

# Read genotype and variant data

```{r}

scripts_folder <- file.path(base_folder,"scripts","s08_consolidate_dataset")
setwd(scripts_folder)

data_folder <- file.path(base_folder,"data","s08_consolidate_dataset")


load(file.path(data_folder,"s02_filter_genotypes.RData"))

ls()

full.vcfr

dim(header.df)
dim(variants.df)

dim(full_dp.mx)
dim(full_gq.mx)
dim(full_gt.mx)

```

# Explore duplicated plate-keys

```{r}

sum(duplicated(participants.df$platekey))

nrow(participants.df) - ncol(full_gt.mx)

duplicates_index <- duplicated(participants.df$platekey)
sum(duplicates_index)

duplicates <- participants.df$platekey[duplicates_index]

duplicated_participants.df <- participants.df[participants.df$platekey %in% duplicates,]

rm(duplicates_index, duplicates, duplicated_participants.df)

```

# Save data

```{r}

save.image(file.path(data_folder,"s03_add_phenotypes.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()

```
