---
title: "Import tables from LabKey"
output: 
  html_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

**Started**: Rofaida, 18Feb2020  
**Last updated**: Rofaida, 13Mar2020

# Summary  

Import selected fields from these four tables:  

- Aggregate VCF stats  
- Participants  
- Sporadic Cancer diagnosis  
- Rare Diseases diagnosis  

Fields and tables were pre-selected manually in exploratory analyis. 

### Field names  

The exact field names are available from the Data Dictionary spreadsheet (link inside GEL-RE):  

- https://re-confluence.gel.zone/pages/viewpage.action?pageId=19630487#MainProgrammeDataReleasev8(28/11/2019)-MainProgrammeReleaseNotev8andDataDictionaryFiles  

### Required configuration  

**.netrc** file was created in the home folder before running the script,  
as described here (link inside GEL-RE):  

- https://re-confluence.gel.zone/display/GERE/LabKey+Examples  

# Start section

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r}
options(width=999)
```

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())  # remove all objects from environment
graphics.off() # remove all plots from environment

# Dont use factors by default
options(stringsAsFactors = F)

# Folders
base_folder <- "/home/rdesoki/re_gecip/inherited_cancer_predisposition/rofaida/Rproject"

scripts_folder <- paste(base_folder,"scripts","s01_import_Labkey_tables",sep="/")
setwd(scripts_folder)

data_folder <- paste(base_folder,"data","s01_import_Labkey_tables",sep="/")
dir.create(data_folder,showWarnings=F)

# Library for connecting to LabKey  
library(Rlabkey)

# Set labkey baseUrl (do not change)  
labkey.setDefaults(baseUrl="http://emb-prod-mre-labkey-01.gel.zone:8080/labkey/")

# folder for data release (do not change)
v8_folder_path="/main-programme/main-programme_v8_2019-11-28" 

```

# aggregate_gvcf_sample_stats table  

Will be used for  

- Link between **plate_key** and **participant_id**  
- Genetically predicted ethnicity  
- Maybe for PC-s and parents platekeys  

```{r}

# Set query: define fields and table (refer to Data Dictionary Main Programme v8)
q <- "SELECT participant_id, 
             platekey, 
             type, 
             sample_type, 
             PC1,
             PC2,
             PC3,
             PC4,
             PC5,
             PC6,
             pred_african_ancestries,
             pred_american_ancestries,
             pred_east_asian_ancestries,
             pred_european_ancestries,
             pred_south_asian_ancestries
      FROM aggregate_gvcf_sample_stats"

# Execute query
aggregated_vcf.df <- labkey.executeSql(
    schemaName="lists", # do not change (the internal name of database)
    colNameOpt="rname", # do not change (makes column names compartible with R)
    maxRows=100000000, # do not change (should exceed the table size in LabKey GUI)
    folderPath=v8_folder_path, # data release
    sql = q # the query name
)

# Check result
dim(aggregated_vcf.df)
str(aggregated_vcf.df)
head(aggregated_vcf.df)

# Clean-up
rm(q)

```

# participant table  

Contains demografics and general descriptive data  

```{r}

# Set query: define fields and table (refer to the )
q <- "SELECT participant_id, 
             duplicated_participant_id, 
             participant_type, 
             biological_relationship_to_proband, 
             participant_phenotypic_sex, 
             year_of_birth, 
             date_of_consent, 
             participant_ethnic_category, 
             father_affected, 
             mother_affected, 
             rare_diseases_family_sk
      FROM participant"

# Execute query
participant.df <- labkey.executeSql(
    schemaName="lists", # do not change (the internal name of database)
    colNameOpt="rname", # do not change (makes column names compartible with R)
    maxRows=100000000, # do not change (should exceed the table size in LabKey GUI)
    folderPath=v8_folder_path, # data release
    sql = q # the query name
)

# Check result
dim(participant.df)
str(participant.df)
head(participant.df)

# Clean-up
rm(q)

```

# cancer_participant_disease table  

Data about sporadic cancer diagnoses  

```{r}

# Set query: define fields and table (refer to the )
q <- "SELECT participant_id, 
             cancer_disease_type, 
             cancer_disease_sub_type
      FROM cancer_participant_disease"

# Execute query
sporadic_cancers.df <- labkey.executeSql(
    schemaName="lists", # do not change (the internal name of database)
    colNameOpt="rname", # do not change (makes column names compartible with R)
    maxRows=100000000, # do not change (should exceed the table size in LabKey GUI)
    folderPath=v8_folder_path, # data release
    sql = q # the query name
)

# Check result
dim(sporadic_cancers.df)
str(sporadic_cancers.df)
head(sporadic_cancers.df)

# Clean-up
rm(q)

```

# cancer_participant_tumour table  

It is imported because it has date of diagnosis for sporadic cancer.  
However, later exploration showed that this column has too much missed values.    
So the date of concent may be used instead.  

```{r}

# Set query: define fields and table (refer to the )
q <- "SELECT participant_id, 
             diagnosis_date, 
      FROM cancer_participant_tumour"

# Execute query
cancer_diagnosis_date.df <- labkey.executeSql(
    schemaName="lists", # do not change (the internal name of database)
    colNameOpt="rname", # do not change (makes column names compartible with R)
    maxRows=100000000, # do not change (should exceed the table size in LabKey GUI)
    folderPath=v8_folder_path, # data release
    sql = q # the query name
)

# Check result
dim(cancer_diagnosis_date.df)
str(cancer_diagnosis_date.df)
head(cancer_diagnosis_date.df)

# Clean-up
rm(q)

```

# rare_diseases_participant_disease table  

Data about RD diagnoses  

```{r}

# Set query: define fields and table (refer to the )
q <- "SELECT participant_id, 
             normalised_disease_group,
             normalised_disease_sub_group,
             normalised_specific_disease,
             normalised_age_of_onset,
             diagnosis_date,
      FROM rare_diseases_participant_disease"

# Execute query
rare_diseases.df <- labkey.executeSql(
    schemaName="lists", # do not change (the internal name of database)
    colNameOpt="rname", # do not change (makes column names compartible with R)
    maxRows=100000000, # do not change (should exceed the table size in LabKey GUI)
    folderPath=v8_folder_path, # data release
    sql = q # the query name
)

# Check result
dim(rare_diseases.df)
str(rare_diseases.df)
head(rare_diseases.df)

# Clean-up
rm(q)

```

# Save  result

```{r}

rm(v8_folder_path)

save.image(file=paste(data_folder,"s01_import_Labkey_tables.RData",sep="/"))

```

# Final section

```{r}

ls()           # list objects in the environment
sessionInfo()  # list loaded libraries etc
Sys.time()     # print time

```
