---
title: "Select cases and controls"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

**Started**:Alexey Larionov, 14Mar20
**Last updated**: Rofaida, 24Mar20  

# Summary  

Select cases and controls basing on the preliminary exploration of data  

# Start section

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r}
options(width=999)
```

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Dont use factors by default
options(stringsAsFactors = F)

# Folders
base_folder <- "/home/rdesoki/re_gecip/inherited_cancer_predisposition/rofaida/Rproject"

# Libraries
library(dplyr)

```

# load data  

```{r}

load(paste(base_folder,"data","s02_explore_Labkey_data","s04_explore_controls.RData", sep="/"))

scripts_folder <- paste(base_folder,"scripts","s03_select_sample",sep="/")


data_folder <- paste(base_folder,"data","s03_select_sample",sep="/")
dir.create(data_folder,showWarnings=F)

ls()

dim(vcf_participants.df)
dim(sporadic_cancers.df)
dim(rare_diseases.df)
dim(unaffected_parents_idd.df)

```

# Select cases and controls

### RD cases

```{r}

rare_diseases.df <- rare_diseases.df %>% 
  filter(normalised_disease_group == "Tumour syndromes")

dim(rare_diseases.df)

```

### Ca cases  

```{r}

breast <- sporadic_cancers.df$cancer_disease_type == "BREAST"
colorectal <- sporadic_cancers.df$cancer_disease_type == "COLORECTAL"
lung <- sporadic_cancers.df$cancer_disease_type == "LUNG"
renal <- sporadic_cancers.df$cancer_disease_type == "RENAL"
sarcoma <- sporadic_cancers.df$cancer_disease_type == "SARCOMA"
endometrial <- sporadic_cancers.df$cancer_disease_type == "ENDOMETRIAL_CARCINOMA"
ovarian <- sporadic_cancers.df$cancer_disease_type == "OVARIAN"
prostate <- sporadic_cancers.df$cancer_disease_type == "PROSTATE"
glioma <- sporadic_cancers.df$cancer_disease_type == "ADULT_GLIOMA"
bladder <- sporadic_cancers.df$cancer_disease_type == "BLADDER"
hepbil <- sporadic_cancers.df$cancer_disease_type == "HEPATOPANCREATOBILIARY"
melanoma <- sporadic_cancers.df$cancer_disease_type == "MALIGNANT_MELANOMA"
upgist <- sporadic_cancers.df$cancer_disease_type == "UPPER_GASTROINTESTINAL"

ca_cases <- breast | colorectal | lung | renal | sarcoma | endometrial | ovarian | 
            prostate | glioma | bladder | hepbil | melanoma | upgist

sum(ca_cases)

sporadic_cancers.df <- sporadic_cancers.df %>% 
  filter(ca_cases)

rm(breast,colorectal,lung,renal,sarcoma,endometrial,ovarian, 
   prostate,glioma,bladder,hepbil,melanoma,upgist,ca_cases)

```

### Check for intersects

Just in case  

```{r}

# Check for intersects
intersect(sporadic_cancers.df$participant_id, rare_diseases.df$participant_id)
intersect(sporadic_cancers.df$participant_id, unaffected_parents_idd.df$participant_id)
intersect(rare_diseases.df$participant_id, unaffected_parents_idd.df$participant_id)

```

# Select cases and controls from **the VCF-participants**  

### Keep "VCF-participants" variables only for cases and controls, add a column with their group info  

```{r}

vcf_ca.df <- vcf_participants.df %>% 
  filter(participant_id %in% sporadic_cancers.df$participant_id) %>% 
  mutate(group="sporadic_ca")
dim(vcf_ca.df)

vcf_rd.df <- vcf_participants.df %>% 
  filter(participant_id %in% rare_diseases.df$participant_id) %>% 
  mutate(group="rd_tumours")
dim(vcf_rd.df)

vcf_controls.df <- vcf_participants.df %>% 
  filter(participant_id %in% unaffected_parents_idd.df$participant_id) %>% 
  mutate(group="control")
dim(vcf_controls.df)

rm(vcf_participants.df)

```

### Merge **VCF-participants** of cases and controls into single data.frame  

```{r}

vcf_cases_controls.df <- rbind(vcf_ca.df, vcf_rd.df, vcf_controls.df)
dim(vcf_cases_controls.df)

sum(duplicated(vcf_cases_controls.df$participant_id))
sum(duplicated(vcf_cases_controls.df$platekey))

rm(vcf_ca.df, vcf_rd.df, vcf_controls.df)

```

# Merge **Ca**, **RD** and **Control** data to one table  

Merging will be done by rbind() : so need to make the same columns in each table  

### Prepare RD

```{r}

str(rare_diseases.df)

rare_diseases.df <- rare_diseases.df %>% 
  mutate(ca_diagnosis_date=as.Date(NA),
         control_relationship_to_proband=NA, 
         control_age_at_consent=NA) %>% 
  select(participant_id, 
         disease_group="normalised_disease_group", 
         disease_sub_group="normalised_disease_sub_group", 
         disease="normalised_specific_disease", 
         rd_age_of_onset="normalised_age_of_onset", 
         rd_diagnosis_date="diagnosis_date", 
         ca_diagnosis_date, 
         control_relationship_to_proband, 
         control_age_at_consent)

str(rare_diseases.df)

```

### Prepare Ca

```{r}

str(sporadic_cancers.df)

sporadic_cancers.df <- sporadic_cancers.df %>% 
  mutate(rd_age_of_onset=NA,
         rd_diagnosis_date=as.Date(NA),
         control_relationship_to_proband=NA, 
         control_age_at_consent=NA) %>% 
  select(participant_id, 
         disease_group="cancer_disease_type", 
         disease_sub_group="cancer_disease_sub_type", 
         disease,
         rd_age_of_onset,
         rd_diagnosis_date,
         ca_diagnosis_date="diagnosis_date",
         control_relationship_to_proband,
         control_age_at_consent) 

str(sporadic_cancers.df)

```

### Prepare Controls

```{r}

str(unaffected_parents_idd.df)

unaffected_parents_idd.df <- unaffected_parents_idd.df %>% 
  mutate(disease_group=NA,
         disease_sub_group=NA,
         disease=NA,
         rd_age_of_onset=NA,
         rd_diagnosis_date=as.Date(NA),
         ca_diagnosis_date=as.Date(NA)) %>% 
  select(participant_id, 
         disease_group, 
         disease_sub_group, 
         disease,
         rd_age_of_onset,
         rd_diagnosis_date,
         ca_diagnosis_date,
         control_relationship_to_proband="biological_relationship_to_proband",
         control_age_at_consent="age_at_consent") 

str(unaffected_parents_idd.df)

```

### Combime RD, Ca and Controls

```{r}

cases_controls.df <- rbind(rare_diseases.df, sporadic_cancers.df, unaffected_parents_idd.df)
dim(cases_controls.df)
str(cases_controls.df)

rm(rare_diseases.df, sporadic_cancers.df, unaffected_parents_idd.df)

```

# Combine cases and controls with VCF data  

Note that some participants have multiple diagnoses  

```{r}

dim(cases_controls.df)
dim(vcf_cases_controls.df)

sum(!vcf_cases_controls.df$participant_id %in% cases_controls.df$participant_id)
sum(!cases_controls.df$participant_id %in% vcf_cases_controls.df$participant_id)

sum(duplicated(vcf_cases_controls.df$participant_id))
sum(duplicated(cases_controls.df$participant_id))

participants.df <- left_join(cases_controls.df, vcf_cases_controls.df, by="participant_id")

dim(participants.df)
colnames(participants.df)

rm(cases_controls.df, vcf_cases_controls.df)

```

# Save  result

```{r}

save.image(paste(data_folder,"s01_select_sample.RData",sep="/"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()

```
