---
title: "Clean data"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

**Started**: Alexey Larionov, 23Mar20  
**Last updated**: Rofaida, 25Mar20

# Summary  

* Explore and clean the data, remove and rearrange columns etc  
* Evaluate age of **consent**, age of **onset** and age of **diagnosis** in Ca and RD  
* Set age for participants:  
    + **Controls** age = age of **consent**  
    + **Sporadic Ca** age = age of **consent**  
    + **RD Tumour Syndromes** age = age of **diagnosis**  

# Start section

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r}
options(width=999)
```

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Dont use factors by default
options(stringsAsFactors = F)

# Folders
base_folder <- "/home/rdesoki/re_gecip/inherited_cancer_predisposition/rofaida/Rproject"

# Libraries
library(dplyr)
library(ggplot2)

```

# load data  

```{r}

# Load data
load(paste(base_folder,"data","s03_select_sample","s01_select_sample.RData", sep="/"))

# Set folders
scripts_folder <- paste(base_folder,"scripts","s03_select_sample",sep="/")
setwd(scripts_folder)

data_folder <- paste(base_folder,"data","s03_select_sample",sep="/")


# Check data
ls()
dim(participants.df)

```

# Remove unnecessary columns

```{r}

# Explore columns
colnames(participants.df)

# Remove columns
participants.df <- participants.df %>% 
  select(- pred_african_ancestries, - pred_american_ancestries, - pred_east_asian_ancestries, 
         - pred_south_asian_ancestries, - father_affected, - mother_affected)

# Explore result
colnames(participants.df)

```

# Check ID duplicates  

ID duplicates are expected because some participants may have multiple diseases  

```{r}

sum(duplicated(participants.df$participant_id))
sum(duplicated(participants.df$platekey))

```

# Remove control's age of consent 

Because its identical to the age of concent available for all participants  

```{r}

# Explore
sum(is.na(participants.df$age_at_consent))
sum(is.na(participants.df$control_age_at_consent))
sum(participants.df$age_at_consent != participants.df$control_age_at_consent, na.rm=T)

# Remove field
participants.df <- participants.df %>% select(-control_age_at_consent)

```

# Keep relationship to proband for controls only

Even for controls it will not be needed any longer ... keeping just in case  

```{r}

# Explore
table(participants.df[,c("participant_type","group")])

sum(is.na(participants.df$biological_relationship_to_proband))
sum(is.na(participants.df$control_relationship_to_proband))

sum(participants.df$biological_relationship_to_proband != 
      participants.df$control_relationship_to_proband, na.rm=T)

table(participants.df[,c("biological_relationship_to_proband","group")])

participants.df %>% 
  group_by(biological_relationship_to_proband) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

participants.df %>% 
  group_by(control_relationship_to_proband) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Remove field
participants.df <- participants.df %>% 
  select(- biological_relationship_to_proband, - participant_type)

```

# Remove unnecessary "type" column

```{r}

# Eexplore
table(participants.df[,c("type","group")])

# Remove column
participants.df <- participants.df %>% 
  select( - type)

```

# Calculate RD and Ca ages, rename and rearrange columns  

Some columns will never be needed, but kept for now just in case

```{r}

# Remind column names
colnames(participants.df)

# Calculate diagnosis ages
participants.df <- participants.df %>% 
  mutate(sex=participant_phenotypic_sex,
         rd_diagnosis_age=as.numeric(format(rd_diagnosis_date,"%Y")) - year_of_birth,
         ca_diagnosis_age=as.numeric(format(ca_diagnosis_date,"%Y")) - year_of_birth) %>% 
  select(platekey, participant_id, group,
         sex, year_of_birth, date_of_consent,age_at_consent,
         rd_age_of_onset,rd_diagnosis_age,ca_diagnosis_age,
         pc1,pc2,pc3,pc4,pc5,pc6,
         disease_group, disease_sub_group, disease,
         sample_type, rare_diseases_family_sk, control_relationship_to_proband,
         pred_european_ancestries, participant_ethnic_category)

# Check result
colnames(participants.df)

```

# Compare Ca **diagnosis** age with the age of **consent**

Ca diagnosis age was not yet easily available for many patients. For the patients, where it has been available, the Ca diagnosis age is well reflected by the age of consent.  

Because the age of consent is available for all Ca patients, it will be used for Ca age at diagnosis.  

```{r}

# Relation between age of Ca diagnosis and enrollment
ggplot(participants.df, aes(x=ca_diagnosis_age,y=age_at_consent)) +
  geom_point() +
  ggtitle("Sporadic cancers")

# Count missed data: diagnosis date had not yet been easily provided many cases
sum(is.na(participants.df[participants.df$group=="sporadic_ca","ca_diagnosis_age"]))
sum(is.na(participants.df[participants.df$group=="sporadic_ca","age_at_consent"]))

# Count age==0: Ca cases with this suspicious age have already been removed
sum(participants.df$group=="sporadic_ca" & participants.df$ca_diagnosis_age==0, na.rm = T)
sum(participants.df$group=="sporadic_ca" & participants.df$age_at_consent==0)

```

# Compare RD **diagnosis** age with the age of **onset** and age of **consent**

We will use age of **diagnosis** for RD age.  However, overall, any of the other "age" fields could be justifiably used for age of participant.  

### Relations between diagnosis, enrollment and onset in all RD patients

Age of **diagnosis** was similar to the age of enrollment (**consent**).  However, the **onset** was often reported much earlier than diagnosis/enrollment. I think that age of **onset** better reports the disease start, so I prefer using it over the other possible ages.  

```{r}

ggplot(participants.df, aes(x=rd_diagnosis_age,y=rd_age_of_onset)) +
  geom_point() +
  ggtitle("RD Tumour Syndromes")

ggplot(participants.df, aes(x=age_at_consent,y=rd_age_of_onset)) +
  geom_point() +
  ggtitle("RD Tumour Syndromes")

ggplot(participants.df, aes(x=rd_diagnosis_age,y=age_at_consent)) +
  geom_point() +
  ggtitle("RD Tumour Syndromes")

```

### Relations between diagnosis, enrollment and onset in familial breast/ovarian Cancer

```{r}

participants.df %>% 
  filter(disease=="Familial breast and or ovarian cancer") %>% 
ggplot(aes(x=rd_diagnosis_age,y=rd_age_of_onset)) +
  geom_point() +
  ggtitle("Familial breast and or ovarian cancer")

participants.df %>% 
  filter(disease=="Familial breast and or ovarian cancer") %>% 
ggplot(aes(x=age_at_consent,y=rd_age_of_onset)) +
  geom_point() +
  ggtitle("Familial breast and or ovarian cancer")

participants.df %>% 
  filter(disease=="Familial breast and or ovarian cancer") %>% 
ggplot(aes(x=rd_diagnosis_age,y=age_at_consent)) +
  geom_point() +
  ggtitle("Familial breast and or ovarian cancer")

```

# Count missed age data

83 RD Tumour Syndromes participants have missed age of **onset** and 89 participants have age of **onset** == 0.  These participants will be removed because many those with onset at 0 years are obvious entry errors (e.g. 11 familial breast cancers at 0 years age).  This will be 83+89=172 of 1,364 RD participants with tumour syndromes.  

```{r}

# Missed data in different age columns in RD  
sum(is.na(participants.df[participants.df$group=="rd_tumours","rd_diagnosis_age"]))
sum(is.na(participants.df[participants.df$group=="rd_tumours","rd_age_of_onset"]))
sum(is.na(participants.df[participants.df$group=="rd_tumours","age_at_consent"]))

# Count age == 0
sum(participants.df$group=="rd_tumours" & participants.df$rd_diagnosis_age==0)
sum(participants.df$group=="rd_tumours" & participants.df$rd_age_of_onset==0, na.rm = T)
sum(participants.df$group=="rd_tumours" & participants.df$age_at_consent==0)

# What RD cases have missed age of onset?
participants.df %>% 
  filter(group=="rd_tumours" & is.na(rd_age_of_onset)) %>% 
  group_by(disease) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count))

# What RD cases have age of onset == 0 ?
participants.df %>% 
  filter(group=="rd_tumours" & rd_age_of_onset==0) %>% 
  group_by(disease) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count))

# What are the total numbers of RD cases ?
participants.df %>% 
  filter(group=="rd_tumours") %>% 
  group_by(disease) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count))

```

# Set age of participants for controls, Ca and RD  

### Make empty column for age

```{r}

# Add new column
participants.df <- participants.df %>% 
  mutate(age=NA)

# Reminder for the counts in "group"
table(participants.df$group)

```

### Age for controls = age of consent

```{r}

# Set controls' age
participants.df[participants.df$group == "control", "age"] <- 
  participants.df[participants.df$group == "control", "age_at_consent"]

# Check result
sum(!is.na(participants.df$age))

```

### Age for sporadic cancers = age of consent

```{r}

# Set Ca age 
participants.df[participants.df$group == "sporadic_ca", "age"] <- 
  participants.df[participants.df$group == "sporadic_ca", "age_at_consent"]

# Check result
sum(!is.na(participants.df$age))
sum(is.na(participants.df$age))

```

### Age for RD Tumour Syndromes = rd_age_of_diagnosis

```{r}

participants.df[participants.df$group == "rd_tumours", "age"] <- 
  participants.df[participants.df$group == "rd_tumours", "rd_diagnosis_age"]

sum(!is.na(participants.df$age))
sum(is.na(participants.df$age))

```

# Remove unnecessary age columns  

```{r}

participants.df <- participants.df %>% 
  select(platekey, participant_id, group, sex, age, 
         pc1,pc2,pc3,pc4,pc5,pc6,
         disease_group, disease_sub_group, disease,
         sample_type, rare_diseases_family_sk, control_relationship_to_proband,
         pred_european_ancestries, participant_ethnic_category)

```

# Save  result

```{r}

save.image(paste(data_folder,"s02_clean_data.RData",sep="/"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()

```
