---
title: "Check relatedness"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

**Started**: Alexey Larionov, 23Mar20  
**Last updated**: Rofaida, 28Mar20  

# Summary  

Make sure that no relatives are selected for analysis, because comparing relatives would violate statisitcal assumption about independence of individuals.  

Relatedness is defined by keenship coefficient > 0.04419417 (3rd degree relatives) as calculated by GEL, see  
- gel_data_resourses/main_programme/aggregated_illumina_vcd/GRCH38/20190228/principal_components_and_relatedness/PCA_readme_20190910.pdf  

The related samples are removed iteratively:  

1) Pairwise related samples are provided by GEL  
2) Related participants in the selected samples are explored:  
- 377 pairs of related samples (657 unique samples) identified, of them  
- 402 are controls, 172 are sporadic cancers and 89 are RD tumour syndromes  
3) 402 related **controls** are removed  
4) Remaining related samples are explored again:  
- 66 pairs of related samples (125 unique samples) identified, of them
- 72 are sporadic cancers and 55 are RD tumour syndromes  
5) 66 related participants (38 **sporadic_ca** and 23 **rd_tumours**) are removed  
6) Remaining samples are explored again, and no related pairs are found  

# Start section

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r}
options(width=999)
```

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Dont use factors by default
options(stringsAsFactors = F)

# Folders
base_folder <- "/home/rdesoki/re_gecip/inherited_cancer_predisposition/rofaida/Rproject"

# Libraries
library(dplyr)

```

# load data  

```{r}

# Participants table
load(paste(base_folder,"data","s03_select_sample","s02_clean_data.RData", sep="/"))

# Folders
scripts_folder <- paste(base_folder,"scripts","s03_select_sample",sep="/")
setwd(scripts_folder)

data_folder <- paste(base_folder,"data","s03_select_sample",sep="/")

# Table with related pairs of participants
related_pairs_all.df <- read.table(paste(scripts_folder,"s03_pairwise_related_samples.txt",sep="/"),
                               sep="\t",quote="",fill=T)

c("sample1","sample2","kinship") -> colnames(related_pairs_all.df)
head(related_pairs_all.df)

# Check the data
ls()
dim(participants.df)
dim(related_pairs_all.df)

```

# Check for duplicates in the pairs of related samples  

There are no  

- pairs entered twice in reverse order  
- duplicated pairs of related samples  

```{r}

# Pairs entered twice in reverse order
direct_pair <- paste(related_pairs_all.df$sample1, related_pairs_all.df$sample2)
reverse_pair <- paste(related_pairs_all.df$sample2, related_pairs_all.df$sample1)
length(intersect(direct_pair,reverse_pair))

# Duplicated pairs
sum(duplicated(direct_pair))
sum(duplicated(related_pairs_all.df$sample1))
sum(duplicated(related_pairs_all.df$sample2))

# Clean-up
rm(direct_pair, reverse_pair)

```

# Explore related pairs in the selected participants  

```{r}

# Make matrix for output
related_pairs_selected.mx <- matrix(nrow=0, ncol=3)
c("sample1","sample2","kinship") -> colnames(related_pairs_selected.mx)

# For each line in related_pairs_all.df
for(i in 1:nrow(related_pairs_all.df)){
  
  # If both related samples are in participants.df
  if(related_pairs_all.df[i,"sample1"] %in% participants.df$platekey & 
     related_pairs_all.df[i,"sample2"] %in% participants.df$platekey){
    
    # Write the related samples pair into the results matrix
    related_pairs_selected.mx <- rbind(related_pairs_selected.mx,related_pairs_all.df[i,])
  }
}

# Check result
dim(related_pairs_selected.mx)

# Convert to data frame
related_pairs_selected.df <- as.data.frame(related_pairs_selected.mx)
str(related_pairs_selected.df)

# Clean-up
rm(i, related_pairs_selected.mx)

```

### Explore related participants  

```{r}

# There is no inverse duplicated pairs
direct_pair <- paste(related_pairs_selected.df$sample1, related_pairs_selected.df$sample2)
reverse_pair <- paste(related_pairs_selected.df$sample2, related_pairs_selected.df$sample1)
length(intersect(direct_pair,reverse_pair))

# There are some individual intersects and duplicates
length(intersect(related_pairs_selected.df$sample1, related_pairs_selected.df$sample2))

sum(duplicated(related_pairs_selected.df$sample1))
sum(duplicated(related_pairs_selected.df$sample2))

all_related <- unique(c(related_pairs_selected.df$sample1, related_pairs_selected.df$sample2))
length(all_related)

# Related samples by the group
participants.df %>% 
  filter(platekey %in% all_related) %>% 
  group_by(group) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Clean-up
rm(direct_pair, reverse_pair)

```

# Remove the related **controls**  

```{r}

related_controls.df <- participants.df %>% 
  filter(platekey %in% all_related & group == "control") %>% 
  select(participant_id)

participants.df <- participants.df %>% 
  filter(!participant_id %in% related_controls.df$participant_id)

rm(all_related, related_controls.df, related_pairs_selected.df)

```

# Select related pairs in the **remaining** participants  

```{r}

# Make matrix for output
related_pairs_selected.mx <- matrix(nrow=0, ncol=3)
c("sample1","sample2","kinship") -> colnames(related_pairs_selected.mx)

# For each line in related_pairs_all.df
for(i in 1:nrow(related_pairs_all.df)){
  
  # If both related samples are in participants.df
  if(related_pairs_all.df[i,"sample1"] %in% participants.df$platekey & 
     related_pairs_all.df[i,"sample2"] %in% participants.df$platekey){
    
    # Write the related samples pair into the results matrix
    related_pairs_selected.mx <- rbind(related_pairs_selected.mx,related_pairs_all.df[i,])
  }
}

# Check result
dim(related_pairs_selected.mx)

# Convert to data frame
related_pairs_selected.df <- as.data.frame(related_pairs_selected.mx)
str(related_pairs_selected.df)

# Clean-up
rm(i, related_pairs_selected.mx)

```

### Explore remaining related pairs  

```{r}

# There are some individual intesects and duplicates
length(intersect(related_pairs_selected.df$sample1, related_pairs_selected.df$sample2))

sum(duplicated(related_pairs_selected.df$sample1))
sum(duplicated(related_pairs_selected.df$sample2))

all_related <- unique(c(related_pairs_selected.df$sample1, related_pairs_selected.df$sample2))
length(all_related)

# Related participants by group
participants.df %>% 
  filter(platekey %in% all_related) %>% 
  group_by(group) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Clean-up
rm(all_related)

```

# Remove related samples in **sporadic_ca** and **rd_tumour_syndromes**

Remove only one sample from the pair  

```{r}

# IDs of the samples to be removed 
related_ca_and_rd.df <- participants.df %>% 
  filter(platekey %in% related_pairs_selected.df$sample2) %>% 
  select(participant_id)

# Groups of the samples to be removed 
participants.df %>% 
  filter(participant_id %in% related_ca_and_rd.df$participant_id) %>% 
  group_by(group) %>% 
  summarise(n())

# Remove the samples
participants.df <- participants.df %>% 
  filter(!participant_id %in% related_ca_and_rd.df$participant_id)

# Clean-up
rm(related_ca_and_rd.df, related_pairs_selected.df)

```

# Check relatedness in remaining participants

No related pairs in the remaining participants  

```{r}

# Make matrix for output
related_pairs_selected.mx <- matrix(nrow=0, ncol=3)
c("sample1","sample2","kinship") -> colnames(related_pairs_selected.mx)

# For each line in related_pairs_all.df
for(i in 1:nrow(related_pairs_all.df)){
  
  # If both related samples are in participants.df
  if(related_pairs_all.df[i,"sample1"] %in% participants.df$platekey & 
     related_pairs_all.df[i,"sample2"] %in% participants.df$platekey){
    
    # Write the related samples pair into the results matrix
    related_pairs_selected.mx <- rbind(related_pairs_selected.mx,related_pairs_all.df[i,])
  }
}

# Check result
dim(related_pairs_selected.mx)

# Clean-up
rm(i, related_pairs_selected.mx, related_pairs_all.df)

```

# Check and save

```{r}

dim(participants.df)

save.image(paste(data_folder,"s04_check_relatedness.RData",sep="/"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()

```
