---
title: "Make samples list"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

**Started**: Alexey Larionov, 26Mar20  
**Last updated**: Rofaida, 05Apr20

# Summary  

The previous steps of analysis:    

- explored available data  
- selected cases with relevant diagnoses and sufficient numbers of samples for analysis  
- selected adult controls that are likely to be non-affected by tumours and not diagnosed with other diseases 
- kept only participants with currently valid consent  
- removed ethnically non-European participants  
- defined age of participants  
- removed potentially blood related cases  

This identified 12,957 participants suitable for downstream statistical analysis  

This script  

- makes a text file that will be used for extracting data from VCF  
- explores dataset to get final numbers of samples in different sub-groups  

# Start section

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r}
options(width=999)
```

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Dont use factors by default
options(stringsAsFactors = F)

# Folders
base_folder <- "/home/rdesoki/re_gecip/inherited_cancer_predisposition/rofaida/Rproject"

# Libraries
library(dplyr)
library(ggplot2)

```

# Load data  

```{r}

# Participants table
load(paste(base_folder,"data","s03_select_sample","s04_check_relatedness.RData", sep="/"))

# Folders
scripts_folder <- paste(base_folder,"scripts","s03_select_sample",sep="/")
setwd(scripts_folder)

data_folder <- paste(base_folder,"data","s03_select_sample",sep="/")
dir.create(data_folder,showWarnings=F)

# Check the data
ls()
dim(participants.df)

```

# Check counts in sub-groups  

```{r}

colnames(participants.df)

participants.df %>% 
  group_by(group) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

participants.df %>% 
  group_by(sex) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

participants.df %>% 
  ggplot(aes(age, fill=group)) +
  geom_density(alpha=0.3) +
  ggtitle("Age in different groups")

participants.df %>% 
  filter(group=="rd_tumours") %>% 
  group_by(disease) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

participants.df %>% 
  filter(group=="sporadic_ca") %>% 
  group_by(disease_group) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

participants.df %>% 
  filter(group=="sporadic_ca") %>% 
  select(participant_id, disease_group, disease) %>%
  group_by(disease_group, disease) %>% 
  summarize(counts=n_distinct(participant_id)) %>% 
  filter(counts > 50) %>% 
  arrange(desc(counts), .by_group=T) %>% 
  print(n=Inf)

```

# Export participants  

Note that there are duplicated platekeys/participant_ids in the file because some participants have several tumours.  
The duplicated platekeys should be removed before the export.  

```{r}

# Check and remove duplicated plate keys
sum(duplicated(participants.df$platekey))
platekeys <- unique(participants.df$platekey)
length(platekeys)

# Write platekeys to the text file
participants_file <- paste(data_folder, "samples.txt", sep="/")

write.table(platekeys, 
            file=participants_file,
            row.names=F, col.names=F,
            quote=F, sep="\t")

# Clean-up
rm(participants_file, platekeys)

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()

```
